Dynamic Stochastic Approximation Monte Carlo

### 2018-05-10 papers/histogram/sad.tex

\paragraph{Initializing Wang-Landau:} We begin the simulation by
defining a minimum important energy and a maximum entropy state. In
traditional Wang-Landau the WL factor begins at 1.0 and the factor is
modified by 2.0 until the simulation is cutoff typically around
$10^{-10}$.  Rane et. al. published an implemenation of
WL-TMMC~\cite{rane2013monte} that suggested using a WL factor between
1.0 and $10^{-2}$ with a cutoff $<10^{-5}$. In our implementation of
WL-TMMC, we set the WL factor to be 1 and the cutoff to be either
$10^{-4}$ or $10^{-10}$. A single sweep is defined as each macrostate
being sampled a number of times in this case 10,000 (although 100 is
often sufficient for small systems). Wang-Landau sets a flatness
criteria~\cite{wang2001determining, wang2001efficient,
hatch2015computational, mahynski2017predicting} for the accumulated
histogram usually between 0.75 to 0.90.  In the original
implementation, it was found to be sufficient that each macrostate is
visited at least once (each histogram bin has one
entry)~\cite{shell2003improved}; however, for other systems this is
rarely sufficient.  The goal of this WL initialization is to prefill
the transition matrix for the next portion of the simulation.
\paragraph{Initializing TMMC:} Prefilling the transition
matrix is necessary because if numerous zeros existed in the collection
matrix, the infinite temperature transition matrix would be
ill-defined.  Sampling over large ranges of density of states would
therefore take an unacceptable length of time to
complete~\cite{shell2003improved, shen2014elucidating}.  The transition
matrix in terms of the collection matrix is given as
follows:
\begin{align}
\widetilde{T}_{\infty}(I\rightarrow J) = \frac{C(I,J)}
{\sum_{K} C(I,K)}
\end{align}
We now run the TMMC portion of the simulation.
%Siderius does this for 20 additional sweeps (does about 8 sweeps for
%WL).  We do this until we reach a specified temperature of interest. We
%have not yet implemented a biasing function to control acceptance rate
%of trial moves but we probably could/should? at some point.
The acceptance probability is written in terms of the infinite temperature
transition matrix.
\begin{align}
  P_{acc}(i\rightarrow j) = \min\bigg[1,\frac{\widetilde{T}_{\infty}(J\rightarrow I)}
  {\widetilde{T}_{\infty}(I\rightarrow J)}\bigg]
\end{align}
This probability is only used to control actual transitions not update
the collection matrices.

@@

\section{SAD Algorithm}
\subsubsection{The theory behind SAD}
The SAD (Dynamic Stochastic Approximation) is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. While for
SAMC the update factor is defined in the original
implementation~\cite{liang2007stochastic} in terms of an arbitrary
tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{t_0}{\max(t_0,t)}
\end{align}
The update factor $\gamma_{t}^{\text{SA}}$ can be thought of as
$\nicefrac{\text{d}S}{\text{d}t}$. This tells us that $t_0$ should have
dimensions of entropy.  We argue that we can write this entropy `term'
geometrically as an area bounded by the $\ln{D(E)}$ i.e. the entropy.  This
gives the following
\begin{align}
\gamma_{t}^{\text{SA}} =
\frac{N_E}{t}\frac{E_{\text{max}}-E_{\text{min}}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state.  The factor of $3$ comes from the ratio of areas bounded
by the $\ln{D(E)}$ and the minimum temperature line.
The update factor has the same $1/t$ behavior as the original SAMC algorithm;
however now every time a new energy is discovered during the course
of the simulation, the update factor experiences a jump.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-n500.pdf}
  \caption{The update factor $\gamma$ for N = 500 versus iteration for
  three different methods: WL, SAMC, and SAD.
    method.}\label{fig:gamma-vs-t}
\end{figure}

@@

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error.pdf}
  \caption{The average entropy error for each method is shown for a filling fraction of 0.30 with
  SAMC followed by TMMC serving as the `golden'
  standard.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_10-N50-entropy-error.pdf}
  \caption{The average entropy error for each method is shown for a filling fraction of 0.10 with
  SAMC followed by TMMC serving as the `golden' standard.}\label{fig:N50-ff0.1-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_20-N50-entropy-error.pdf}
  \caption{The average entropy error is shown for a filling fraction of 0.20 with TMI acting as the
  `golden' standard.}\label{fig:N50-ff0.2-avg-error}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error.pdf}
  \caption{The average entropy error is shown for a filling fraction of 0.30 with TMI acting as the
  `golden' standard.}\label{fig:n500-ff0.2}
\end{figure}

\subsubsection{A periodic system with 10\% filling fraction and $(N = 50)$}
In order to determine how the various methods scale with size,
simulations are run for systems with different particle numbers. A filling
fraction $\text{ff} = 0.10$ is chosen which yields $N = 50$ particles
in the cubical system.  These particles are randomly inserted into the
system and random moves are proposed (with a translation scale
$\delta_0 = 0.05$). The boundaries in $x$ and $y$ are periodic while
the $z$-axis is not periodic.

We use the average entropy error vs iterations~\jpsays{we should
explain how we do this.} as a metric to compare simulation runtimes and
overall convergence.  Simulations that take more iterations to reach
the same entropy error as a different method take longer to converge.
This behavior is seen in Fig.~\ref{fig:N50-ff0.1-avg-error}.

@@

Figure~\ref{fig:N50-ff0.2-avg-error} shows a system consisting of 50 atoms
that are placed in a periodic simple cubic unit cell with lattice constant
chosen to give a filling fraction of 0.20.  These particles are
randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

@@

Figure~\ref{fig:N50-ff0.3-avg-error} shows a system consisting of 50 atoms
that are placed in a periodic simple cubic unit cell with lattice constant
chosen to give a filling fraction of 0.30.  These particles are
randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

@@

A system is configured with dimensions $x = y = z = 19\sigma$.  A
filling fraction $\text{ff} = 0.30$ is chosen which yields $N = 500$
particles in a cubical system.  These particles are randomly inserted
into the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05$). The boundaries in $x$ and $y$ are periodic while
the $z$-axis is not periodic.

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$ for the 
Wang-Landau algorithm decreases much more rapidly than $\nicefrac{1}{\sqrt{t}}$.
It can be observed that all the methods from this family struggle to converge if
the update factor decreases faster than $\nicefrac{10^5}{\sqrt{t}}$. Fig.~\ref{fig:n500-ff0.2}
shows both WL and SAMC with inappropriate choices of $t_0$ i.e. less than 
$\nicefrac{10^5}{\sqrt{t}}$ failing to converge in a reasonable amount of time.
Thus, the update factor $\gamma$, shown as a function of simulation moves
in Fig.~\ref{fig:gamma-vs-t}, proves useful in determining which methods will 
converge in a reasonable amount of time.
We compare each method to a `golden' standard which we define to be an ideal choice
of SAMC for $10^11$ moves followed by TMMC.  This gives an unbiased comparison with
our developed method SAD.

@@

We introduced a new algorithm SAD (which is Stochastic Approximation
with the modification factor dynamically chosen) that effectively
samples the entire energy space for a variety of system sizes.  Also,
SAD proves effective in comparison with other Monte-Carlo methods at
converging to the correct density of states.  The ease of
implementation and the absence of user-defined parameters makes SAD an
ideal algorithm.

~\jpsays{We need about a paragraph here that says exactly how much
better SAD performed than other methods and or how easy it was to
implement (Likely best added once simulations run a little longer).}

We note that SAD should be tested on additional systems, and the
accuracy of the method should be examined.  Also, a rigorous
mathematical justification for our choice of the update factor is
necessary to ensure the most optimal convergence of the algorithm.

>>>>

\paragraph{Initializing Wang-Landau:} We begin the simulation by
defining a minimum important energy and a maximum entropy state. In
traditional Wang-Landau the WL factor begins at 1.0 and the factor is
modified by 2.0 until the simulation is cutoff typically around
$10^{-10}$.  Rane et. al. published an implemenation of
WL-TMMC~\cite{rane2013monte} that suggested using a WL factor between
1.0 and $10^{-2}$ with a cutoff $<10^{-5}$. In our implementation of
WL-TMMC, we set the WL factor to be 1 and the cutoff to be either
$10^{-4}$ or $10^{-10}$. A single sweep is defined as each macrostate
being sampled a number of times in this case 10,000 (although 100 is
often sufficient for small systems). Wang-Landau sets a flatness
criteria~\cite{wang2001determining, wang2001efficient,
hatch2015computational, mahynski2017predicting} for the accumulated
histogram usually between 0.75 to 0.90.  In the original
implementation, it was found to be sufficient that each macrostate is
visited at least once (each histogram bin has one
entry)~\cite{shell2003improved}; however, for other systems this is
rarely sufficient.  The goal of this WL initialization is to prefill
the transition matrix for the next portion of the simulation.
\paragraph{Initializing TMMC:} Prefilling the transition
matrix is necessary because if numerous zeros existed in the collection
matrix, the infinite temperature transition matrix would be
ill-defined.  Sampling over large ranges of density of states would
therefore take an unacceptable length of time to
complete~\cite{shell2003improved, shen2014elucidating}.  The transition
matrix in terms of the collection matrix is given as
follows:
\begin{align}
\widetilde{T}_{\infty}(I\rightarrow J) = \frac{C(I,J)}
{\sum_{K} C(I,K)}
\end{align}
We now run the TMMC portion of the simulation.
%Siderius does this for 20 additional sweeps (does about 8 sweeps for
%WL).  We do this until we reach a defined number of iterations. We
%have not yet implemented a biasing function to control acceptance rate
%of trial moves but we could? at some point.
The acceptance probability is written in terms of the infinite temperature
transition matrix.
\begin{align}
  P_{acc}(i\rightarrow j) = \min\bigg[1,\frac{\widetilde{T}_{\infty}(J\rightarrow I)}
  {\widetilde{T}_{\infty}(I\rightarrow J)}\bigg]
\end{align}
This probability is only used to control actual transitions not update
the collection matrices.

@@

\section{SAD Algorithm}
\subsubsection{The theory behind SAD}
The SAD (Dynamic Stochastic Approximation) is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. While for
SAMC the update factor is defined in the original
implementation~\cite{liang2007stochastic} in terms of an arbitrary
tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{t_0}{\max(t_0,t)}
\end{align}
The update factor $\gamma_{t}^{\text{SA}}$ can be thought of as
$\nicefrac{\text{d}S}{\text{d}t}$. This tells us that $t_0$ should have
dimensions of entropy.  We argue that we can write this entropy `term'
geometrically as an area bounded by the $\ln{D(E)}$ i.e. the entropy.  This
gives the following
\begin{align}
\gamma_{t}^{\text{SA}} =
\frac{N_E}{t}\frac{E_{\text{max}}-E_{\text{min}}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state.  The factor of $3$ comes from the ratio of areas bounded
by the $\ln{D(E)}$ and the minimum temperature line.
The update factor has the same $1/t$ behavior as the original SAMC algorithm;
however now every time a new energy is discovered during the course
of the simulation, the update factor experiences a jump. This behavior allows
SAD to dynamically prevent the update factor from decreasing too rapidly.

@@

\subsubsection{High and Low energy racheting}
Since SAD explores all energy states it is imperative that the method have
a reasonable idea about where the minimum important energy and maximum
entropy state are located in energy space. Since these are not user input
parameters like for methods suchas WL-TMMC, the simulation is responsible
for determining and updating these energy locations. In order to prevent,
the method from experiencing errors near the energy extrema, we employ a 
high energy and a low energy to define the range over which the energy 
histogram is made flat. The high energy is the highest ever value that the
maximum entropy state has taken while the low energy is the lowest ever value
that the minimum important energy has taken. Together these two energies serve
as a kind of ratcheting effect to determine the energy landscape boundaries.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-n500}
  \caption{The update factor $\gamma$ for $N=500$ versus iteration for three
    different methods: WL, SAMC, and SAD.
    method.}\label{fig:gamma-vs-t}
\end{figure}

@@

We use the average entropy error vs moves a metric to compare 
simulation runtimes and overall convergence. The overall accuracy
is determined by examining the fractional error of a particular method to
a `golden' standard. For each simulation, the `golden' standard is
SAMC for the best choice of $t_0$ for $10^{11}$ moves followed by TMMC.  
\begin{align}
\epsilon_\text{avg} = 1 - \bigg\lvert\frac{S_\text{method}}{S_\text{golden}}\bigg\rvert
\end{align}
Simulations that take more iterations to reach the same entropy error as a 
different method take longer to converge. This behavior is seen in 
Fig.~\ref{fig:N50-ff0.1-avg-error}.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_10-N50-entropy-error.pdf}
  \caption{The average entropy error for each method is shown for a filling fraction of 0.10 with
  SAMC followed by TMMC serving as the `golden' standard.}\label{fig:N50-ff0.1-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_20-N50-entropy-error.pdf}
  \caption{The average entropy error is shown for a filling fraction of 0.20 with TMI acting as the
  `golden' standard.}\label{fig:N50-ff0.2-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error.pdf}
  \caption{The average entropy error for each method is shown for a filling fraction of 0.30 with
  SAMC followed by TMMC serving as the `golden'
  standard.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error.pdf}
  \caption{The average entropy error is shown for a filling fraction of 0.30 with TMI acting as the
  `golden' standard.}\label{fig:n500-ff0.2}
\end{figure}

\subsubsection{A periodic system with 10\% filling fraction and $(N = 50)$}
In order to determine how the various methods scale with size,
simulations are run for systems with different particle numbers. 
Figure~\ref{fig:N50-ff0.1-avg-error} shows the average entropy vs moves for
a system consisting of 50 atoms that are placed in a periodic simple cubic unit
cell with lattice constant chosen to give a filling fraction of 0.10.  These particles
are randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

@@

Figure~\ref{fig:N50-ff0.2-avg-error} shows the average entropy vs moves for
a system consisting of 50 atoms that are placed in a periodic simple cubic unit
cell with lattice constant chosen to give a filling fraction of 0.20.  These particles
are randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

@@

Figure~\ref{fig:N50-ff0.3-avg-error} shows the average entropy vs moves for
a system consisting of 50 atoms that are placed in a periodic simple cubic unit
cell with lattice constant chosen to give a filling fraction of 0.30.  These particles
are randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

@@

A system is configured with dimensions $x = y = z = 19\sigma$.  A
filling fraction $\text{ff} = 0.30$ is chosen which yields $N = 500$
particles in a cubical system.  These particles are randomly inserted
into the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05$).

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$
for the Wang-Landau algorithm decreases much more rapidly than
$\nicefrac{1}{\sqrt{t}}$. It can be observed that all the methods from
this family struggle to converge if the update factor decreases faster
than $\nicefrac{10^5}{\sqrt{t}}$. Fig.~\ref{fig:n500-ff0.2} shows both
WL and SAMC with inappropriate choices of $t_0$ i.e. less than
$\nicefrac{10^5}{\sqrt{t}}$ failing to converge in a reasonable amount
of time. Thus, the update factor $\gamma$, shown as a function of
simulation moves in Fig.~\ref{fig:gamma-vs-t}, proves useful in
determining which methods will converge in a reasonable amount of time.
We compare each method to a `golden' standard which we define to be an
ideal choice of SAMC for $10^{11}$ moves followed by TMMC.  This gives an
unbiased comparison with our developed method SAD. The dynamic version
of SAMC which we call SAD performs as well as SAMC (assuming the user
can correctly identify an appropriate choice of $t_0$).  SAD
outperforms WL and WL-TMMC for the same given number of moves. The
absence of user defined parameters makes SAD an exciting alternative to
methods where user experience or successive implementation iteration
are needed to achieve reasonable results.

@@

We have introduced a new algorithm SAD (which is Stochastic Approximation
with the modification factor dynamically chosen) that effectively
samples the entire energy space for a variety of system sizes.  Also,
SAD proves effective in comparison with other Monte-Carlo methods at
converging to the correct density of states.  The ease of
implementation and the absence of user-defined parameters makes SAD an
ideal algorithm. SAD converges as quickly as SAMC (in the case of an ideally
chosen $t_0$ for our simulations. SAD doesn't suffer from the weakness of SAMC,
i.e. the user choosing parameters that could lead to the simulation not
converging in a reasonable amount of time. SAD also converges more quickly
than TMMC (the reader will remember that SAMC followed by TMMC is what each
method is compared against) and much more quickly than WL-TMMC.

SAD decreases $\gamma$ in a controlled way that prevents the algorithm
from taking too long to converge (as in the case of SAMC with a poor choice
of $t_0$) or failing to converge such as is the case with WL for complex
systems. A major advantage is that the user does not need to determine a
priori how quickly the update factor $\gamma$ should be decreased as the
method SAD dynamically determines this.

We note that SAD should be tested on additional systems, and the
convergence of the method should be examined.  Also, a rigorous
mathematical justification for our choice of the update factor is
necessary to ensure the most optimal convergence of the algorithm.
